---
title: "Homework 2"
author: "Yuki Joyama"
date: "2023-09-26"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
options(scipen=999) # display digits of an integer without scientific notation
```

# Problem 1
```{r, echo=F, message=F}
library(tidyverse)
```

Import and clean the data in pols-month.csv

```{r}
pols_month_df =
  read_csv("./fivethirtyeight_datasets/pols-month.csv") |> # import pols_month.csv file
  separate(col=mon, into=c("year", "month", "day")) |> # separate mon variable into year, month and day
  mutate(
    year = as.integer(year), # change the class of year, month, day to integer
    month = as.integer(month),
    day = as.integer(day),
    month = case_match(
    month,
    01 ~ "jan",
    02 ~ "feb",
    03 ~ "mar",
    04 ~ "apr",
    05 ~ "may",
    06 ~ "jun",
    07 ~ "jul",
    08 ~ "aug",
    09 ~ "sep",
    10 ~ "oct",
    11 ~ "nov",
    12 ~ "dec"
    ), # replace month number with month name
    president = case_when(
      prez_gop == 1 ~ "gop",
      prez_dem == 1 ~ "dem"
    ) # make a new variable president with values gop and dem
  ) |> 
  select(-c("prez_dem", "prez_gop", "day")) # remove prez_dem, prez_gop, and day

pols_month_df
```

Import and clean the data in snp.csv

```{r}
snp_df =
  read_csv("./fivethirtyeight_datasets/snp.csv") |> # import snp.csv file
  separate(col = date, into = c("month", "day", "year")) |> # separate date variable into year, month and day
  mutate(
    year = as.integer(year), # change the class of year, month, day to integer
    month = as.integer(month),
    day = as.integer(day),
    year = case_when(
      year < 23 ~ 2000 + year,
      year >=23 ~ 1900 + year
    ) # if year < 23, add 2000; if year >= 23, add 1900
  ) |> 
  arrange(year, month) |> # arrange order according to year and month
  mutate(
    month = case_match(
    month,
    01 ~ "jan",
    02 ~ "feb",
    03 ~ "mar",
    04 ~ "apr",
    05 ~ "may",
    06 ~ "jun",
    07 ~ "jul",
    08 ~ "aug",
    09 ~ "sep",
    10 ~ "oct",
    11 ~ "nov",
    12 ~ "dec"
    ) # replace month number with month name
  ) |> 
  relocate(year, month) |> # change position of columns year and month
  select(-"day") # remove day

snp_df
```

Import and clean the data in unemployment.csv

```{r}
unemployment_df =
  read_csv("./fivethirtyeight_datasets/unemployment.csv") |> # import unemployment.csv file
  janitor::clean_names() |> # clean the variables name
  pivot_longer(
    cols = c("jan","feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec"),
    names_to = "month",
    values_to = "unemployment_rate"
  ) # change the dataset from wide to long

unemployment_df
```

Now, let's merge snp into pols, and merge unemployment data into the result.

```{r}
fivethrityeight_df =
  left_join(pols_month_df, snp_df, by = c("year", "month")) |> # merge snp into pols,
  left_join(unemployment_df, by = c("year", "month")) |> # merge unemployment data into the result
  relocate("year", "month", "president", "close", "unemployment_rate") # bring key variables to the front

fivethrityeight_df
```

After the cleaning process, pols_month_df has 9 variables including year, month, president (whether the president was republican or democratic), etc; snp_df has three variables including year, month, and close (the closing values of the S&P stock index at the associated time); unemployment_df has three variables including year, month, and unemployment_rate (percentage of unemployment at the associated time).  
Merging these three datasets gives the resulting dataset fivethirtyeight_df. This dataset has 11 variables (year, month, president, close, unemployment_rate, etc and 822 rows. The year in this dataset ranges from 1947 to 2015.

# Problem 2

```{r}
# import dataset from each sheet in excel file
mister_df =
  readxl::read_excel("Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N549") |> # data of Mr. Trash Wheel
  janitor::clean_names()

professor_df =
  readxl::read_excel("Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = "A2:M96") |> # data of Professor Trash Wheel
  janitor::clean_names()

gwynnda_df =
  readxl::read_excel("Trash Wheel Collection Data.xlsx", sheet = "Gwynnda Trash Wheel", range = "A2:K108") |> # data of Gwynnda Trash Wheel
  janitor::clean_names()
```

*Homes Powered - Each ton of trash equates to on average 500 kilowatts of electricity. An average household will use 30 kilowatts per day.

Following the above note, I will calculate homes_powered variable for every dumpster data and update it in each dataset. I will also add an additional variable which specifies the name of Trash Wheel.

```{r}
mister_df = mister_df |> 
  mutate(
    homes_powered = (weight_tons * 500) / 30,
    name = "mister"
  ) |> 
  relocate("name") # bring the name to the front

professor_df = professor_df |> 
  mutate(
    homes_powered = (weight_tons * 500) / 30,
    name = "professor"
  ) |> 
  relocate("name") # bring the name to the front

gwynnda_df = gwynnda_df |> 
  mutate(
    homes_powered = (weight_tons * 500) / 30,
    name = "gwynnda"
  ) |> 
  relocate("name") # bring the name to the front
```

Now, let's combine the three datasets.

```{r}
# change year variable in mister_df from chr to int
mister_df = mister_df |> 
  mutate(year = as.integer(year))

# combine datasets
trash_wheel_df =
  bind_rows(mister_df, professor_df, gwynnda_df)
```

The resulting dataset trash_wheel_df has 747 observations and 16 variables such as name (the name of the trash wheel), month, year, dumpster (the dumpster number), weight_tons (amount of total litter), variables indicating litter type, etc.

The total weight of trash collected by Professor Trash Wheel was `r sum(pull(professor_df, weight_tons))` tons.

```{r}
# filter gwynnda data by year and month 
gwynnda_cig_july2021 = gwynnda_df |> 
  filter(year == 2021 & month == "July") 

gwynnda_cig_july2021 = sum(pull(gwynnda_cig_july2021, cigarette_butts))
```
The total number of cigarette butts collected by Gwynnda in July of 2021 was `r gwynnda_cig_july2021`.






